Detecting and quantifying parasite-induced host mortality from intensity data: method comparisons and limitations

### 2015-08-15 docs/parasite_mortality_manuscript.tex
Made final figure and manuscript updates before revised submission

% \begingroup
\setlength{\parindent}{0in}
\renewcommand{\arraystretch}{2}
% \renewcommand{\tabcolsep}{0}
%\renewcommand{\tabrowsep}{1}
\begin{tabular}{p{3cm} p{10cm}}

\textbf{Title:} & Detecting and quantifying parasite-induced host mortality from intensity data: method comparisons and limitations \\

\end{tabular}

\begin{tabular}{p{3cm} p{10cm}}

\textbf{Authors:} & Mark Q. Wilber, Sara B. Weinstein and Cheryl J. Briggs \\

\end{tabular}

\begin{tabular}{p{3cm} p{10cm}}

\textbf{Affiliations:} &
1) Department of Ecology, Evolution and Marine Biology, University of California,
Santa Barbara, Santa Barbara California, United States of America \\

\end{tabular}

\begin{tabular}{p{3cm} p{10cm}}

\textbf{Corresponding \newline author:} &
Mark Wilber \newline
University of California, Santa Barbara \newline
Department of Ecology, Evolution, and Marine Biology \newline
2111 Noble Hall \newline
Santa Barbara, CA 93106 \newline
505-321-5381 \newline
mark.wilber@lifesci.ucsb.edu \newline

\end{tabular}

\textbf{Note}: Supplementary data associated with this article

% \renewcommand{\arraystretch}{1}
% \begin{tabular}{l l}

% \textbf{Words - Abstract:} & 293 \\
% \textbf{Words - Main Text:} & 4988 \\
% \textbf{References:} & 50 \\
% \textbf{Figures:} & 5 \\
% \textbf{Tables:} & 1

% \end{tabular}

@@

In terms of the host survival function, the Likelihood Method gave unbiased estimates of survival function parameter $a$ when sample sizes were large, however as sample size decreased these estimates became severely biased (Fig. \ref{fig:question2}, Supplementary Fig. S7 - S9) The Adjei Method produced
biased estimates of the host survival function across all sample sizes, with consistently greater bias for steeper survival functions and higher mean parasite loads. (Fig. \ref{fig:question2}, Supplementary Fig. S7-S9).

>>>>

\noindent
\textbf{Title:} Detecting and quantifying parasite-induced host mortality from intensity data: method comparisons and limitations

\bigskip
\noindent
\textbf{Authors:} Mark Q. Wilber, Sara B. Weinstein and Cheryl J. Briggs

\bigskip
\noindent
\textbf{Affiliations:} 1) Department of Ecology, Evolution and Marine Biology, University of California,
Santa Barbara, Santa Barbara California, United States of America

\bigskip
\noindent
\textbf{Corresponding author:}\\Mark Q. Wilber\\University of California, Santa Barbara\\
Department of Ecology, Evolution, and Marine Biology\\
2111 Noble Hall\\
Santa Barbara, CA 93106\\
mark.wilber@lifesci.ucsb.edu

\bigskip\noindent
\textbf{Note}: Supplementary data associated with this article

@@

In terms of the host survival function, the Likelihood Method gave unbiased estimates of survival function parameter $a$ when sample sizes were large, however as sample size decreased these estimates became severely biased (Fig. \ref{fig:question2}, Supplementary Fig. S7-S9) The Adjei Method produced
biased estimates of the host survival function across all sample sizes, with consistently greater bias for steeper survival functions and higher mean parasite loads. (Fig. \ref{fig:question2}, Supplementary Fig. S7-S9).

### 2015-08-15 docs/supplementary_material.tex
A few grammar changes.
Made final figure and manuscript updates before revised submission

We provide an implementation and unit tests of the Crofton Method in  Supplementary Information 4.  Figure S\ref{fig:crof_test} visually shows that our implementation of the Crofton Method agrees with results previously published by \cite{Crofton1971a}.

@@

The Adjei Method for estimating PIHM has two steps \citep{Adjei1986}.  The first
step is to estimate the parameters of the pre-mortality host-parasite
distribution using the Crofton Method (see \emph{SI} 1).  The three parameters estimated are the
total number of hosts before mortality $N_p$,  the mean number of parasites per
host before mortality $\mu_p$, and the aggregation of parasites before
mortality given by the parameter $k_p$ from a negative binomial distribution.
When $k_p$ is small, parasites are highly aggregated among hosts and when
$k_p$ is large parasites are more evenly distributed across hosts \citep{Wilson2002}.  The implementation of the Crofton Method has been discussed at length elsewhere \citep[e.g.][ and in \emph{SI} 1]{Royce1990,Lester1984} and we provide a tested implementation of the method in \emph{SI} 4.

@@

\begin{align*}
    \log{\frac{p}{1 - p}} &= a_1 - b \log{x_1}\\
    - \log{\frac{p}{1 - p}} &= a_2 - b \log{x_2}\\
    0 &= a_1 - a_2 - b\log{x_1} + b\log{x_2}\\
    a_2 - a_1 &= b(\log{x_2} - \log{x_1})
\end{align*}
If $a_2 - a_1 = 1$, than the change in log parasite intensity at which $p$ percentage of hosts survive is $1 / b$.

>>>>

We provide an implementation and unit tests of the Crofton Method in  Supplementary Data S3.  Figure S10 visually shows that our implementation of the Crofton Method agrees with results previously published by \cite{Crofton1971a}.

@@

The Adjei Method for estimating PIHM has two steps \citep{Adjei1986}.  The first
step is to estimate the parameters of the pre-mortality host-parasite
distribution using the Crofton Method (see Supplementary Data S1).  The three parameters estimated are the
total number of hosts before mortality $N_p$,  the mean number of parasites per
host before mortality $\mu_p$, and the aggregation of parasites before
mortality given by the parameter $k_p$ from a negative binomial distribution.
When $k_p$ is small, parasites are highly aggregated among hosts and when
$k_p$ is large parasites are more evenly distributed across hosts \citep{Wilson2002}.  The implementation of the Crofton Method has been discussed at length elsewhere \citep[e.g.][ and in Supplementary Data S1]{Royce1990,Lester1984} and we provide a tested implementation of the method in Supplementary Data S3.

@@

\begin{align*}
    \log{\frac{p}{1 - p}} &= a_1 - b \log{x_1}\\
    - \log{\frac{p}{1 - p}} &= a_2 - b \log{x_2}\\
    0 &= a_1 - a_2 - b\log{x_1} + b\log{x_2}\\
    a_2 - a_1 &= b(\log{x_2} - \log{x_1})
\end{align*}
If $a_2 - a_1 = 1$, then the change in log parasite intensity at which $p$ percentage of hosts survive is $1 / b$.
